Density-dependent selection and the limits of relative fitness

### 2018-03-17 manuscript.tex
latest

In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. However, there are reasons to doubt that this setup is ecologically valid. Many absolute fitness models of density-regulated growth exhibit a strong connection between selection and the ability to keep growing at higher densities \citep{fisher_1930,kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). Density also changes in response to selection (even if only transiently) in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}. This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps, and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}. It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

@@

In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

However, the preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$. All of the models of density-dependent selection mentioned above are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. In complex age-structured models, dependence on density is assumed to come from a ``critical age-group'', and this age group's density responds to selection in much the same way as found in the above models without age structure \citep[Chap. 4]{charlesworth_1994} (e.g. adaptive evolution proceeds in the direction of increasing equilibrium density in the critical age-group). However, in the constant-$s$ approximation, reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.

Although similar ideas featured prominently in early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology. For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types. As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

>>>>

In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. However, there are reasons to doubt that this setup is ecologically valid. Many absolute fitness models of density-regulated growth exhibit a strong connection between selection and the ability to keep growing at higher densities \citep{fisher_1930,kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}. This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}). Density also changes in response to selection in the Lotka-Volterra competition model except in special cases (even if only transiently; \citealt{smouse_1976,mallet_2012}). This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps, and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}. It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

@@

In light of this difficulty, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

However, the preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$. Every model of density-regulated growth mentioned above is missing an important aspect of selection which has ramifications for the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}. Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles. Although similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}),

much of the later reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957} rather than the issues surrounding the constant-$s$ approximation.

For instance, \cite{kimura1969natural} simply assumed that $N$ should remain constant

as a requirement and then derived variants of the logistic model consistent with this. \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.

In complex age-structured models, dependence on density is assumed to come from a ``critical age-group'', and this age group's density responds to selection in much the same way as found in the above models without age structure \citep[Chap. 4]{charlesworth_1994} (e.g. adaptive evolution proceeds in the direction of increasing equilibrium density in the critical age-group).

  As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.
